Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2022-10-31 src/tex/ms.tex

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO Virgo Kagra Gravitational Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any; we
present inferences for the recent GWTC-3 transient catalog that suggest that the
stable mass transfer efficiency parameter may vary with primary black hole mass.  
\end{abstract}

@@

We apply the method described in section \ref{sec:method} to the first GW event
GW150914. The parameters involved in the analysis and the range we allow for the
inference are tabulated in table \ref{tab:parameters}. The spin of BHs formed in
isolated binaries depends strongly on the angular momentum transport in massive
stars which is still highly uncertain (see however \citet{Fuller2019,
Bavera2020}), therefore we only consider the two component masses in the
observable space in both the root-finding stage and the MCMC stage. For
progenitor parameters, we characterize each guess with five parameters: the two
component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period
$t_{\rm orb}$, the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that
the progenitor formation redshift is not obtained through the root-finding
process because the formation redshift does not intrinsically affect the
evolution of the binary. Instead, we compute the time it takes for the binary to
evolve from formation to the merger and add this time to the lookback time when
a GW event is observed and limit the total time to merger to be less than
$13.7\,\rm{Gyr}$. Using the total lookback time, we can compute the redshift of
the binary at ZAMS formation following the Planck 2018 cosmological model
\citep{Planck2018}. The main benefit of post-processing the redshift in this way
is that we do not rely on a particular assumption of star formation rate (SFR)
distribution when we solve for the posterior distribution in redshift. This
means we can put the prior on formation redshift (or in general,
metallicity-redshfit distribution) in by reweighting the posterior distribution
according to a particular SFR model, and therefore do not have to rerun the
inference every time we change the SFR model. This is especially helpful since
SFR models are still highly uncertain but strongly impact the local merger rates
of BBHs \citep{Broekgaarden2021}. For hyperparameters, we choose parameters that
affect binary evolution in \texttt{COSMIC} most significantly for massive stars,
including the common envelope efficiency $\alpha$, the fraction of mass accreted
onto the accretor during stable mass transfer $f_{\rm acc}$, the critical mass
ratio which determines whether Roche-overflow mass transfer proceeds stably or
produces a common envelope when the donor star is on the Hertzsprung gap $q_{\rm
crit,3}$, and the root-mean-square of the Maxwellian distribution employed for
CO natal kicks, $\sigma$.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level,
we perform the same analysis for all events in GWTC-3. Plotted in figure
\ref{fig:GWTC-3_f_acc_mass} is the posterior density in the $m_{1,{\rm
GW}}-f_{\rm acc}$ space for most of the events in GWTC-3. Each contour
represents the $68\%$ credible interval of the posterior density for that
particular event. There is a suggestive trend showing that $f_{\rm acc}$ could
increase as the mass of the progenitor increases. Such a trend implies the
stable mass transfer phase of a less massive binary would be preferentially
non-conservative, with more conservative mass transfer in more massive binary
systems.

@@

% Rate analysis is now just counting, like what's done in phenomenological
% model, instead of forward modelling.
Working event-by-event as we do permits post-processing application of
arbitrarily complicated models of star formation and metalliticy evolution by
re-weighting the samples we obtain \citep[e.g.][]{VanSon2022}. Once we have
pulled the GW event posterior back from the observable space to the evolutionary
parameter space, we can apply the same hierarchical inference methods currently
used to infer the compact binary population to our progenitor population.  In
particular, we can use our progenitor population to infer the formation rate of
compact binary population progenitor systems over cosmic time without needing to
make artifical and simplistic assumptions about the delay time distribution or
the metallicity-specific star formation rate
\citep{Vitale2019,Ng2021,VanSon2022}.  Comparing the formation rate of
progenitors to the star formation rate provides an avenue to check our
understanding of binary evolution and the relative rate of other formation
channels.

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space for
the first time, there are a number of improvements that can be implemented in
future studies. In this study we use only \texttt{COSMIC} as our evolutionary
function, which by design cannot explain all the events in GWTC-3. For example,
event with either of the component mass larger than the lower edge of PISN gap
such as GW190521 cannot be explained by \texttt{COSMIC}. Alternative channels
such as dynamical formation might be needed to explain some subset of the events
in GWTC-3 \citep{Zevin2021}. As the sensitivity of GW detectors increases, we
expect to see more and more events that are unusual in some way. Therefore,
having a self-consistence population synthesis code that contains multiple
formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating"
GW events posterior samples, highlighting the capability of our method and
motivating the community to build the next generation population synthesis code
that can work with our method. To avoid cluttering of focus, we discuss the
physical implication of the result presented in this work under our specific
assumption (i.e. using \texttt{COSMIC} as our evolutionary function) in a
companion paper.

@@

To summarize, we propose a novel method to recover the posterior samples in the
evolutionary parameter space for each GW event. We point out hyperparameters in
the usual population synthesis simulation context are not actually parameters
related to the population, but parameters about the evolutionary function. This
means the binary evolution functions can be constrained on an individual event
basis. We "back propagate" the posterior in the observable space to the
evolutionary parameter space, thus allowing us to study hyperparameters and
theirs correlations with progenitor parameters in a data-driven manner. Our
method makes less assumptions than the traditional forward modelling approach,
which often fix the hyperparameters across the entire population. Since we are
not limited to the fixed hyperparameters assumptions, we can explore the
behavior of the hyperparameters across the population much more efficiently.
While our work lays down a data analysis pathway to understand the population of
GW events, no physics can be learned without a comprehensive physical model. We
hope this letter will motivate the community to build the next generation
population synthesis codes that have the following properties: first, they
should have explicit control over the random variables so marginalizing over
random variables can be done more precisely; second, they should be as
automatically differentiable as much as possible so exploring the evolutionary
parameter space is efficient. Combining this work and the next generation of
population synthesis codes, we can explore the full parameter space of binary
evolution models with the next-generation GW detectors network in the near
future.

\section{Acknowledgement}
The authors are grateful for discussions with the CCA Gravitational Wave
Astronomy Group. The Flatiron Institute is supported by the Simons Foundation.

>>>>

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO--Virgo Gravitational-Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any; we
present inferences for the recent GWTC-3 transient catalog that suggest that the
stable mass transfer efficiency parameter may vary with primary black hole mass.  
\end{abstract}

@@

We apply the method described in section \ref{sec:method} to the first GW event GW150914 \citep{GW150914}.
We use the \texttt{Overall\_posterior} posterior samples for GW150914,
publicly available from the Gravitational wave Open Science Center \citep{LIGOScientific:2019lzm}. 
The parameters involved in the analysis and the range we allow for the inference are tabulated in table \ref{tab:parameters}.
The spin of BHs formed in isolated binaries depends strongly on the angular
momentum transport in massive stars which is still highly uncertain (see however \citet{Fuller2019, Bavera2020}),
therefore we only consider the two component masses in the observable space in both the
root-finding stage and the MCMC stage. For progenitor parameters, we characterize each guess with
five parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period $t_{\rm orb}$,
the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does not intrinsically affect
the evolution of the binary. Instead, we compute the time it takes for the binary to evolve from
formation to the merger and add this time to the lookback time when a GW event is observed and limit the total
time to merger to be less than $13.7\,\rm{Gyr}$.
Using the total lookback time, we can compute the redshift of the binary at ZAMS formation following the
Planck 2018 cosmological model \citep{Planck2018}.
The main benefit of post-processing the redshift in this way is that we do not rely on a particular assumption
of star formation rate (SFR) distribution when we solve for the posterior distribution in redshift.
This means we can put the prior on formation redshift (or in general, metallicity-redshfit distribution) in
by reweighting the posterior distribution according to a particular SFR model,
and therefore do not have to rerun the inference every time we change the SFR model. This is especially helpful
since SFR models are still highly uncertain but strongly impact the local merger rates of BBHs \citep[e.g. ][]{Broekgaarden2021}.
For hyperparameters, we choose parameters that strongly affect
binary evolution in \texttt{COSMIC} for massive stars, including the common envelope
efficiency $\alpha$, the fraction of mass accreted onto the accretor during stable mass transfer $f_{\rm acc}$,
the critical mass ratio which determines whether Roche-overflow mass transfer
proceeds stably or produces a common envelope when the donor star is on the Hertzsprung gap
$q_{\rm crit,3}$, and the root-mean-square of the Maxwellian distribution
employed for CO natal kicks, $\sigma$. For systems which begin a common envelope 
with a Hertzsprung-gap donor, we always assume the pessimistic outcome that the two 
stars merge since the donor has likely not formed a strong core-envelope boundary 
\citep{Ivanova2004, Belczynski2008}.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level,
we perform the same analysis for all events in GWTC-3. For events announced in
GWTC-1, we use \texttt{Overall\_posterior},
\texttt{PublicationSamples},\texttt{IMRPhenomXPHM\_comoving} and
\texttt{C01:Mixed} posterior samples, respectively. All the data are publicly
available from the Gravitational wave Open Science Center
\citep{LIGOScientific:2019lzm}. Plotted in figure \ref{fig:GWTC-3_f_acc_mass} is
the posterior density in the $m_{1,{\rm GW}}-f_{\rm acc}$ space for
most\footnote{\textbf{We excluded events for which our reconstructed posterior has KL
divergence greater than 0.1 nats relative to the GWTC-3 posterior; see below.}}
of the events in GWTC-3. Each contour represents the $68\%$ credible interval of
the posterior density for that particular event. There is a suggestive trend
showing that $f_{\rm acc}$ could increase as the mass of the progenitor
increases. Such a trend implies the stable mass transfer phase of a less massive
binary would be preferentially non-conservative, with more conservative mass
transfer in more massive binary systems.

@@

% Rate analysis is now just counting, like what's done in phenomenological
% model, instead of forward modelling.
Working event-by-event as we do permits post-processing application of
arbitrarily complicated models of star formation and metalliticy evolution by
re-weighting the samples we obtain \citep[e.g.][]{VanSon2022}. Once we have
pulled the GW event posterior back from the observable space to the evolutionary
parameter space, \textbf{and have accounted for selection effects}, we can apply
the same hierarchical inference methods currently used to infer the compact
binary population to our progenitor population.  In particular, we can use our
progenitor population to infer the formation rate of compact binary population
progenitor systems over cosmic time without needing to make artifical and
simplistic assumptions about the delay time distribution or the
metallicity-specific star formation rate \citep{Vitale2019,Ng2021,VanSon2022}.
Comparing the formation rate of progenitors to the star formation rate provides
an avenue to check our understanding of binary evolution and the relative rate
of other formation channels.

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space, there are a number of improvements that can be implemented in future studies.
In this study we use only \texttt{COSMIC} as our evolutionary function, which by design cannot explain all the events in GWTC-3.
For example, events with either of the component mass larger than the lower edge of PISN gap such as GW190521 cannot be explained by \texttt{COSMIC}.
Alternative channels such as dynamical formation might be needed to explain some subset of the events in GWTC-3 \citep{Zevin2021}.
As the sensitivity of GW detectors increases, we expect to see more and more events that are unusual in some way.
Therefore, having a self-consistence population synthesis code that contains multiple formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating" GW-event posterior samples,
highlighting the capability of our method and motivating the benefits of building population synthesis codes that can work seamlessly with our method.
To avoid cluttering of focus, we discuss the physical implications of the results presented in this work under our specific assumptions (i.e. using \texttt{COSMIC} as our evolutionary function) in a future companion paper.

@@

To summarize, we propose a method to recover the posterior samples in the evolutionary parameter space for each GW event.
We point out hyperparameters in the usual population synthesis simulation context are not actually parameters related to the population,
but parameters about the evolutionary function.
This means the binary evolution functions can be constrained on an individual event basis.
We "back propagate" the posterior in the observable space to the evolutionary parameter space,
thus allowing us to study hyperparameters and theirs correlations with progenitor parameters in a data-driven manner.
Our method makes less assumptions than the traditional forward modelling approach,
which often fix the hyperparameters across the entire population.
Since we are not limited to the fixed hyperparameters assumptions, we can explore the behavior of the hyperparameters across the population much more efficiently.
While our work lays down a data analysis pathway to understand the population of GW events,
no physics can be learned without a comprehensive physical model.
We hope this letter will motivate the construction of next-generation population synthesis codes that have the following properties:
first, they should retain explicit control over the random variables so marginalizing over random variables can be done precisely and
second, they should be as automatically differentiable as possible so that exploring the evolutionary parameter space is efficient.
By combining the methods presented in this work and future, differentiable population synthesis codes,
we can explore the full parameter space of binary evolution models with future GW data.

\section{Acknowledgement}
The authors are grateful for discussions with the CCA Gravitational Wave Astronomy Group.
The Flatiron Institute is supported by the Simons Foundation.  We thank M. Zevin for helpful comments and suggestions on this manuscript.
This research has made use of data or software obtained from the Gravitational Wave Open Science Center (gw-openscience.org),
a service of LIGO Laboratory, the LIGO Scientific Collaboration, the Virgo Collaboration, and KAGRA.
LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as
the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS),
and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector.
Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO),
by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef,
with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain.
The construction and operation of KAGRA are funded by Ministry of Education, Culture, Sports, Science and Technology (MEXT),
and Japan Society for the Promotion of Science (JSPS), National Research Foundation (NRF) and Ministry of Science and ICT (MSIT) in Korea,
Academia Sinica (AS) and the Ministry of Science and Technology (MoST) in Taiwan.

Data produced in the process of this work can be found at \cite{wong2022-data}.
